---
title: "Johns Hopkins COVID-19 Data"
author: "Jason Horne"
date: "6/13/2022"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

# Background

As part of the response to the global COVID-19 pandemic, research institutions
around the world have collected and published data online related to the spread
of the virus. Johns Hopkins provides one such set of data, publishing US and
global case and death counts to a GitHub repository that is free to access. 

We will investigate these data to answer some questions, including:
1. How well do the COVID-19 cases correlate to deaths?
2. Have death rates remained the same through the spread of the Delta and
Omicron variants?
3. How has the state of North Carolina fared in the pandemic, compared to the
rest of the country?


# Load and Standardize
First, we load the data by URL and perform some simple transformations to make
them easier to work with.

```{r import-data}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_global.csv",
                "time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in, file_names)

global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])
```

The data come as a single row per political unit (country, state/province,
county, etc) with columns for each date.

Let's use pivot_longer to transform
the data such that we have a unique row per political unit per date. This will
dramatically increase the number of rows in the data, but likewise dramatically
decrease the number of columns in the data.

```{r tidy1}
global_cases <- global_cases %>% 
  pivot_longer(
    cols = -c(`Province/State`, `Country/Region`, Lat, Long),
    names_to = "date",
    values_to = "cases"
  ) %>% 
  select(-c (Lat, Long))

global_deaths <- global_deaths %>% 
  pivot_longer(
    cols = -c(`Province/State`, `Country/Region`, Lat, Long),
    names_to = "date",
    values_to = "deaths"
  ) %>% 
  select(-c (Lat, Long))

US_cases <- US_cases %>% 
  pivot_longer(
    cols = -(UID:Combined_Key),
    names_to = "date",
    values_to = "cases"
  ) %>% 
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c (Lat, Long_))

US_deaths <- US_deaths %>% 
  pivot_longer(
    cols = -c(UID:Combined_Key, Population),
    names_to = "date",
    values_to = "deaths"
  ) %>% 
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c (Lat, Long_))
```

Note that global and US data are each separated into datasets
for the number of cases and the number of deaths. Let's use the join functions
to put those together. 

After this step, there should be one row per political unit per date, with each
row having both the case count and the death count.


``` {r tidy2}
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>%
  mutate(date = mdy(date)) %>%
  filter(cases > 0)

US <- US_cases %>%
  full_join(US_deaths)
```

Next, we'll add a Combined_Key field to the global data. Then we'll load a
separate dataset with more country-specific information and join it to our
global data by state/province and country.

```{r tidy3}
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))


global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID,FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths,
         Population, Combined_Key)
```


## Summary of Input Data
Let's take a quick look at the data we have so far.

```{r summary}
summary(US)
summary(global)
```

We note that there are rows in the US data that have a negative count of cases
and deaths. Let's look more closely.

```{r negativecases}
US %>% filter(cases < 0)
US %>% filter(deaths < 0)
```

It seems there are two rows from South Carolina with both negative case and
death counts. Why could that be?

For now, let's just remove those rows as possible errors, though we should
later investigate to see if they're attempting to model a correction in previous
numbers, or similar.

```{r filternegative}
US <- US %>% filter(cases >= 0)
summary(US)
```

Now all of our data have nonnegative case and death counts.

# Normalize Counts per Capita
To compare US states with large populations against those with small
populations, we need to look at a per capita case and death rate. We will do
this by calculating the number of cases and deaths per million.

First, we summarize all of the counties within a state. This US_by_state dataset
will have one row per state per date.
```{r visualize1}
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths,
         deaths_per_mill, Population) %>%
  ungroup()

US_by_state
```
Our US_totals will summarize across the states. The resulting dataset will have
one row per date, showing the number of total cases and deaths nationwide by
that date.
```{r visualize2}
US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Country_Region, date, cases, deaths,
         deaths_per_mill, Population) %>%
  ungroup()

US_totals
```

Let's quickly visualize the overall number of cases and deaths within the USA
and the state of North Carolina. We'll use a logarithmic scale so early
ups and downs aren't obscured by wilder variation later in the pandemic.

```{r visualize3}
US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID-19 in USA", y = NULL)

state <- "North Carolina"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID-19 in ", state), y = NULL)
```

These are interesting, but they really obscure the surges from the Delta and
Omicron variants that occurred in late 2021 and early 2022. Let's take off the
logarithmic scale and regenerate the graphs.

```{r visualize4}
US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  #scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID-19 in USA", y = NULL)

state <- "North Carolina"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  #scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID-19 in ", state), y = NULL)
```

These clearly show the COVID-19 case count in the USA and in North Carolina
increasing through early 2021, at which point vaccinations became available.
The total case counts started to flatten until the Delta variant became
prominent starting in the fall of 2021, then flattened again briefly before an
enormous spike from the Omicron variant in the beginning of 2022.

We can visualize this even more clearly if we only look at the incremental
number of new cases and deaths per date (rather than the totals of each). To do
this, we'll add new_cases and new_deaths columns to the datasets, which we
can then plot.

```{r analyze}
US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))


US_totals %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID-19 in USA", y = NULL)

US_by_state %>%
  filter(Province_State == state) %>%
  filter(new_cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID-19 in ", state), y = NULL)
```

These plots clearly show the spikes for the Delta and Omicron variants, though
the death counts are buried at the bottom of the graph since their numbers are
so much lower than the case counts.  Let's look at just the death numbers to 
see if we find the same spikes.

```{r analyze2}
US_totals %>%
  ggplot(aes(x = date, y = new_deaths)) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID-19 in USA", y = NULL)

US_by_state %>%
  filter(Province_State == state) %>%
  filter(new_cases > 0) %>%
  ggplot(aes(x = date, y = new_deaths)) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID-19 in ", state), y = NULL)
```

We can clearly see the spikes for the Delta and Omicron variants in these
plots, but the Omicron death counts are roughly the same as those from Delta,
and not much larger as the Omicron case count was compared to the Delta case
count.

This supports the common assertion that, although Omicron was much more
widespread, it was a 'safer' form of the virus that did not result in as
many per capita deaths as the earlier variants.

# North Carolina vs National
How did the state of North Carolina do in comparison to the whole country?

We'll create a new tibble, joining NC data against the US data by date. We'll
rename the columns so we can distinguish the NC values from the US values, and
then plot them together.

```{r nc_vs_us_1}
NC_vs_US <- US_by_state %>% 
  filter(Province_State == "North Carolina") %>% 
  rename(nc_cases = cases, nc_deaths = deaths, 
         nc_deaths_per_mill = deaths_per_mill, nc_new_cases = new_cases,
         nc_new_deaths = new_deaths, nc_population = Population) %>%
  select(date:nc_new_deaths) %>% 
  full_join(US_totals) %>% 
  select(-c(Country_Region)) %>% 
  rename(us_cases = cases, us_deaths = deaths, 
         us_deaths_per_mill = deaths_per_mill, us_population = Population, 
         us_new_cases = new_cases, us_new_deaths = new_deaths)

NC_vs_US %>%
  ggplot(aes(x = date, y = nc_deaths_per_mill)) +
  geom_line(aes(color = "nc_deaths_per_mill")) +
  geom_point(aes(color = "nc_deaths_per_mill")) +
  geom_line(aes(y = us_deaths_per_mill, color = "us_deaths_per_mill")) +
  geom_point(aes(y = us_deaths_per_mill, color = "us_deaths_per_mill")) +
  #scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID-19 Deaths per Million in North Carolina vs USA", y = NULL)
```

This seems to indicate that North Carolina has had a lower 
number of deaths per million residents than the overall nation throughout
the pandemic.

# Modeling NC vs US

Let's generate a linear model between NC and US deaths per million residents to
see how well the NC rate predicts the US rate.

```{r nc_vs_us_2}
mod_nc_us <- lm(us_deaths_per_mill ~ nc_deaths_per_mill, data = NC_vs_US)
summary(mod_nc_us)
```

This shows a very high correlation between the two, with a p-value less than
2.2e-16.

Let's plot the model to see the relationship visually.

```{r nc_vs_us_3}
NC_vs_US_w_pred <- NC_vs_US %>% mutate(pred = predict(mod_nc_us))
NC_vs_US_w_pred %>% 
  ggplot() + 
  geom_point(
    aes(x = nc_deaths_per_mill, y = us_deaths_per_mill), color = "blue") + 
  geom_point(aes(x = nc_deaths_per_mill, y = pred), color="red")
```
Indeed, we can see a very strong linear relationship between the number of
deaths per million in North Carolina versus in the USA as a whole. The plot
winds very tightly back and forth across the linear model.


```{r analyze3, include=FALSE}
# extra code used in class, but not for my own analysis

US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases), 
            population = max(Population),
            cases_per_thou = cases * 1000 / population,
            deaths_per_thou = deaths * 1000 / population) %>%
  filter(cases > 0, population > 0)

US_state_totals %>% slice_min(deaths_per_thou, n = 10)
US_state_totals %>% slice_max(deaths_per_thou, n = 10)
US_state_totals %>% slice_max(cases_per_thou, n = 10)
```

```{r model, include=FALSE}
# extra code used in class, but not for my own analysis

mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
# unneeded
#x_grid <- seq(1,400)
#new_df <- tibble(cases_per_thou = x_grid)
US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))

US_tot_w_pred %>% ggplot() + geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")

```

# Potential Bias
There are many possible sources of bias in this analysis, including:

1. COVID-19 numbers are reported via agencies within each state, who may have
different procedures for collecting the data within that state. This could
obscure the actual relationship between COVID-19 cases and deaths across states.
2. We excluded data points with zero or negative numbers of cases and death
rates. Why were those there? Were they an attempt to correct errors in data
previously reported?  This should be investigated.
3. As home testing has become more available, fewer people are testing through
health care facilities or laboratories. Positive results from home tests are
not necessarily reported to relevant state agencies. Therefore, the actual
case numbers are likely higher than are reflected in the data for the time
periods when home tests are widely used.


# Session Information
```{r sessioninfo}
sessionInfo()
```